---
title: Implementing Efron's double Poisson distribution in Stan
authors:
- admin
date: '2023-09-12'
codefolding_show: 'show'
slug: double-poisson-in-Stan
categories: []
draft: true
tags:
  - simulation
  - distribution-theory
header:
  caption: ''
  image: ''
---

$$
\def\Pr{{\text{Pr}}}
\def\E{{\text{E}}}
\def\Var{{\text{Var}}}
\def\Cov{{\text{Cov}}}
\def\bm{\mathbf}
\def\bs{\boldsymbol}
$$
For a project I am working on, we are using [Stan](https://mc-stan.org/) to fit generalized random effects location-scale models to a bunch of count data. We're interested in using the double-Poisson distribution, as described by [Efron (1986)](https://doi.org/10.2307/2289002), but it's turning out to be a bit tricky. The double-Poisson distribution is not implemented in Stan, so we've had to write our own distribution function. That's fine and not particularly difficult. What's more challenging is that also we need to also write a Stan function to generate random samples from the double-Poisson, so that we can generate posterior predictive checks. In this post, I'll walk through the implementation of the custom distribution functions needed to use the double-Poisson in Stan. 

The incredible [`gamlss.dist` package](https://cran.r-project.org/package=gamlss.dist) provides a full set of distributional functions for the double-Poisson distribution, including a sampler. Thus, I can validate my Stan functions against the functions from `gamlss.dist`. 

```{r setup, warning = FALSE, message = FALSE}
library(tidyverse)
library(patchwork)   # composing figures
library(gamlss.dist) # DPO distribution functions
library(rstan)       # Stan interface to R
library(brms)        # fitting generalized linear models
library(bayesplot)   # Examine fitted models
library(loo)         # Model fit measures
rstan_options(auto_write = TRUE)
```

## The double-Poisson 

The double-Poisson distribution is a discrete distribution for non-negative counts, with support $\mathcal{S}_X = \{0, 1, 2, 3, ...\}$. 
It is an interesting distribution because it admits for both over- and under-dispersion relative to the Poisson distribution, whereas most of the conventional alternatives such as the negative binomial distribution or Poisson-normal mixture distribution allow only for over-dispersion. 
The mean-variance relationship of the double-Poisson is approximately constant; for $X \sim DPO(\mu, \theta)$,  $\text{E}(X) \approx \mu$ and $\text{Var}(X) \approx \mu / \theta$, so that the double-Poisson distribution approximately satisfies the assumptions of a quasi-Poisson generalized linear model (although not quite exactly so). 

[Efron (1986)](https://doi.org/10.2307/2289002) gives the following expression for the density of the double-Poisson distribution with mean $\mu$ and inverse-disperson $\theta$:
$$
f(x | \mu, \theta) = \frac{\theta^{1/2} e^{-\theta \mu}}{c(\mu,\theta)} \left(\frac{e^{-x} x^x}{x!}\right) \left(\frac{e \mu}{x}\right)^{\theta x},
$$
where $c(\mu,\theta)$ is a scaling constant to ensure that the density sums to one, which is closely approximated by 
$$
c(\mu, \theta) \approx 1 + \frac{1 - \theta}{12 \mu \theta}\left(1 + \frac{1}{\mu \theta}\right).
$$
We then have
$$
\ln f(x | \mu, \theta) = \frac{1}{2} \ln \theta - \theta \mu - \ln c(\mu, \theta) + x (\theta + \theta \ln \mu - 1) + (1 - \theta) x \ln(x) - \ln \left(x!\right),
$$
where $x \ln (x)$ is evaluated as 0.

# Log of the probability mass function

For purposes of using this distribution in Stan, it's sufficient to provide the log of the probability mass function up to a constant---there's no need to normalize it to sum to one. Thus, we can ignore the $c(\mu, \theta)$ term above. Here's a Stan function implementing the lpmf:

```{r lpmf}
stancode_lpmf <- "
functions {
  real dpo_lpmf(int X, real mu, real theta) {
    real ans;
    real A = inv(2) * log(theta) - theta * mu;
    if (X == 0)
      ans = A;
    else
      ans = A + X * (theta * (1 + log(mu)) - 1) - lgamma(X + 1) + (1 - theta) * X * log(X);
    return ans;
  }
}
"
```

To check that this is accurate, I'll compare the Stan function to the corresponding function from `gamlss.dist` for a couple of different parameter values and for $x = 0,...,100$. If my function is accurate, the calculated log-probabilities should differ by a constant value for each set of parameters.

```{r check-lpmf, cache = TRUE}
writeLines(stancode_lpmf, "DPO-lpmf.stan")
expose_stan_functions("DPO-lpmf.stan")

test_lpmf <- 
  expand_grid(
    mu = c(2, 5, 10, 20),
    theta = c(0.1, 0.2, 0.5, 1, 2, 5, 10),
    X = 0:100
  ) %>%
  mutate(
    gamlss_lpmf = dDPO(x = X, mu = mu, sigma = 1 / theta, log = TRUE),
    my_lpmf = pmap_dbl(.l = list(X = X, mu = mu, theta = theta), .f = dpo_lpmf),
    diff = my_lpmf - gamlss_lpmf
  )
```

```{r, fig.width = 7, fig.height = 4, fig.retina = 2}
ggplot(test_lpmf, aes(factor(theta), diff, color = factor(theta))) + 
  geom_boxplot() + 
  facet_wrap( ~ mu, labeller = "label_both", ncol = 2) + 
  theme_minimal() + 
  labs(x = "theta") + 
  theme(legend.position = "none")
```

Checks out. Onward!

# Cumulative distribution function

I'll next implement a function to evaluate the cumulative distriution function over a range of values. This is an expensive calculation, but it can be improved a little bit by noting the relationship between sequential values of the probability mass function. Letting $d = \exp \left(\theta + \theta \ln \mu - 1 \right)$, observe that
$$
\begin{aligned}
f(0 | \mu, \theta) &= \frac{\theta^{1/2} e^{-\theta \mu}}{c(\mu,\theta)} \\
f(1 | \mu, \theta) &= f(0 | \mu, \theta) \times d \\
f(x | \mu, \theta) &= f(x - 1 | \mu, \theta) \times d \times \frac{\exp\left[(1 - \theta)(x - 1)\left(\ln(x) - \ln(x - 1)\right) \right]}{x^\theta}
\end{aligned}
$$
where the last expression holds for $x \geq 2$.

The function below computes the un-normalized cumulative distribution function (i.e., without the scaling constant $c(\mu, \theta)$) over the range $x = 0,...,m$ as follows: 

* Compute $f(x | \mu, \theta)$ for $x = 0,1,2,...$, without the scaling constant $c(\mu, \theta)$
* Take $F(0 | \mu, \theta) = f(0 | \mu, \theta)$ and accumulating $F(x | \mu, \theta) = F(x - 1 | \mu, \theta) + f(x | \mu, \theta)$ for $x = 0,...,m$.
* Checking if $f(x | \mu, \theta) / F(x | \mu, \theta)$ is small (less than $10^{-8}$), in which case accumulation stops at the value $n$.
* The normalized cumulative distribution function will then be $F(x | \mu, \theta) / F(n | \mu, \theta)$.

```{r cdf}
stancode_cdf <- " 
functions {
  vector dpo_cdf(int max_value, real mu, real theta) {
    real d = exp(theta * (1 + log(mu)) - 1);
    real prob;
    int n = max_value + 1;
    vector[n] cdf;
    cdf[1] = sqrt(theta) * exp(-mu * theta);
    prob = cdf[1] * d;
    cdf[2] = cdf[1] + prob;
    for (i in 2:max_value) {
      prob = prob * d * exp((1 - theta) * (i - 1) * (log(i) - log(i - 1))) / (i^theta);
      cdf[i + 1] = cdf[i] + prob;
      if (prob / cdf[i + 1] < 1e-8) {
        n = i + 1;
        break;
      }
    }
    return cdf / cdf[n];
  }
}
"
```

To check that this is accurate, I'll again compare the Stan function to the corresponding function from `gamlss.dist`. If my function is accurate, the computed cdf values should be proportional to the cdf calculated from `gamlss.dist::pDPO()` and the ratio should be very close to 1.

```{r check-cdf, cache = TRUE}
writeLines(stancode_cdf, "DPO-cdf.stan")
expose_stan_functions("DPO-cdf.stan")

test_cdf <- 
  expand_grid(
    mu = c(2, 5, 10, 20),
    theta = c(0.1, 0.2, 0.5, 1, 2, 5, 10),
  ) %>%
  mutate(
    my_cdf = pmap(.l = list(max_value = 100, mu = mu, theta = theta), .f = dpo_cdf)
  ) %>%
  unnest(my_cdf) %>%
  filter(!is.nan(my_cdf)) %>%
  group_by(mu, theta) %>%
  mutate(
    q = row_number() - 1L,
    gamlss_cdf = pDPO(q = q, mu = mu, sigma = 1 / theta),
    ratio = my_cdf / gamlss_cdf
  )
```

```{r, fig.width = 7, fig.height = 4, fig.retina = 2}

ggplot(test_cdf, aes(factor(theta), ratio, color = factor(theta))) + 
  geom_boxplot() + 
  facet_wrap( ~ mu, labeller = "label_both", ncol = 2) + 
  theme_minimal() + 
  labs(x = "theta") + 
  theme(legend.position = "none")

```

Still on track here (although you might wonder, would I be sharing this post if I couldn't get the function working?).

# Quantile function and sampler

The main other thing we need is a function for generating random samples from the double-Poisson. The `gamlss.dist` package has the function `rDPO()` for this purpose. It's implemented using the standard inversion method, by calculating quantiles of the double-Poisson corresponding to a random sample from a uniform distribution. I'll take the same approach.

The function below calculates quantiles by finding the minimum value of $q \geq 0$ such that $F(q + 1 | \mu, \theta) \geq p$ for a specified probability $p \in [0, 1]$, It is vectorized over $p$ and solves for $q$ by starting with the smallest $p$ and continuing through the largest value.

```{r quantile}
stancode_quantile <- " 
functions {
  vector dpo_cdf(real mu, real theta, int max_value) {
    real d = exp(theta * (1 + log(mu)) - 1);
    real prob;
    int n = max_value + 1;
    vector[n] cdf;
    cdf[1] = sqrt(theta) * exp(-mu * theta);
    prob = cdf[1] * d;
    cdf[2] = cdf[1] + prob;
    for (i in 2:max_value) {
      prob = prob * d * exp((1 - theta) * (i - 1) * (log(i) - log(i - 1))) / (i^theta);
      cdf[i + 1] = cdf[i] + prob;
      if (prob / cdf[i + 1] < 1e-8) {
        n = i + 1;
        break;
      }
    }
    return cdf / cdf[n];
  }
  array[] int dpo_quantile(vector p, real mu, real theta, int max_value) {
    int N = rows(p);
    array[N] int qs;
    array[N] int indices = sort_indices_asc(p);
    vector[max_value + 1] cdf_vec = dpo_cdf(max_value, mu, theta);
    int j = 0;
    for (i in indices) {
      while (cdf_vec[j + 1] < p[i]) {
        j += 1;
      }
      qs[i] = j;
    }
    return qs;
  }
}
"
```

If my quantile function is accurate, it should match the value computed from `gamlss.dist::qDPO()` exactly.

```{r check-quantile, cache = TRUE}
writeLines(stancode_quantile, "DPO-quantile.stan")
expose_stan_functions("DPO-quantile.stan")

test_quantile <- 
  expand_grid(
    mu = c(2, 5, 10, 20),
    theta = c(0.1, 0.2, 0.5, 1, 2, 5, 10),
  ) %>%
  mutate(
    p = map(1:n(), ~ runif(100)),
    my_q = pmap(.l = list(p = p, mu = mu, theta = theta, max_value = 5000), .f = dpo_quantile),
    gamlss_q = pmap(.l = list(p = p, mu = mu, sigma = 1 / theta), .f = qDPO)
  ) %>%
  unnest(c(p, my_q, gamlss_q)) %>%
  mutate(
    diff = my_q - gamlss_q
  )
```

```{r, fig.width = 7, fig.height = 4, fig.retina = 2}

ggplot(test_quantile, aes(factor(theta), diff, color = factor(theta))) + 
  geom_boxplot() + 
  facet_wrap( ~ mu, labeller = "label_both", ncol = 2) + 
  theme_minimal() + 
  labs(x = "theta") + 
  theme(legend.position = "none")

```

Phew, still got it!

The last piece of the puzzle is to write a sampler by generating random points from a uniform distribution, then computing the double-Poisson quantiles of these random points. I will implement this two ways: first with an argument for the number of random variates to generate and then, more simply, to generate a single random variate:

```{r rng}
stancode_qr <- "
functions {
  real dpo_lpmf(int X, real mu, real theta, int max_value) {
    real ans;
    real A = inv(2) * log(theta) - theta * mu;
    if (X == 0)
      ans = A;
    else
      ans = A + X * (theta * (1 + log(mu)) - 1) - lgamma(X + 1) + (1 - theta) * X * log(X);
    return ans;
  }
  vector dpo_cdf(real mu, real theta, int max_value) {
    real d = exp(theta * (1 + log(mu)) - 1);
    real prob;
    int n = max_value + 1;
    vector[n] cdf;
    cdf[1] = sqrt(theta) * exp(-mu * theta);
    prob = cdf[1] * d;
    cdf[2] = cdf[1] + prob;
    for (i in 2:max_value) {
      prob = prob * d * exp((1 - theta) * (i - 1) * (log(i) - log(i - 1))) / (i^theta);
      cdf[i + 1] = cdf[i] + prob;
      if (prob / cdf[i + 1] < 1e-8) {
        n = i + 1;
        break;
      }
    }
    return cdf / cdf[n];
  }
  array[] int dpo_quantile(vector p, real mu, real theta, int max_value) {
    int N = rows(p);
    array[N] int qs;
    array[N] int indices = sort_indices_asc(p);
    vector[max_value + 1] cdf_vec = dpo_cdf(max_value, mu, theta);
    int j = 0;
    for (i in indices) {
      while (cdf_vec[j + 1] < p[i]) {
        j += 1;
      }
      qs[i] = j;
    }
    return qs;
  }
  array[] int dpo_sampler(int n, real mu, real theta, int max_value) {
    vector[n] p;
    for (i in 1:n) {
      p[i] = uniform_rng(0,1);
    }
    array[n] int x = dpo_quantile(p, mu, theta, max_value);
    return x;
  }
  int dpo_rng(real mu, real theta, int max_value) {
    real p = uniform_rng(0,1);
    int x = dpo_quantile(p, mu, theta, max_value);
    return x;
  }
}
"
```

To check this function, I'll generate some large samples from the double-Poisson with a few different parameter sets. If the sampler is working properly, the empirical cumulative distribution should line up closely with the cumulative distribution computed using `gamlss.dist::pDPO()`.

```{r check-rng, cache = TRUE}
writeLines(stancode_qr, "DPO-rng.stan")
expose_stan_functions("DPO-rng.stan")

test_rng <- 
  expand_grid(
    mu = c(2, 5, 10, 20),
    theta = c(0.1, 0.2, 0.5, 1, 2, 5, 10),
  ) %>%
  mutate(
    x = pmap(.l = list(n = 10000, mu = mu, theta = theta, max_value = 5000), .f = dpo_sampler),
    tb = map(x, ~ as.data.frame(table(.x)))
  ) %>%
  dplyr::select(-x) %>%
  group_by(mu, theta) %>%
  unnest(tb) %>%
  mutate(
    .x = as.integer(levels(.x))[.x],
    Freq_cum = cumsum(Freq),
    gamlss_F = 10000 * pDPO(q = .x, mu = mu, sigma = 1 / theta)
  )
```

```{r, fig.width = 7, fig.height = 10, fig.retina = 2}
ggplot(test_rng, aes(gamlss_F, Freq_cum, color = factor(theta))) + 
  geom_abline(slope = 1, color = "blue", linetype = "dashed") + 
  geom_point() + geom_line() +  
  facet_grid(theta ~ mu, labeller = "label_both") + 
  theme_minimal() + 
  labs(x = "Theoretical cdf (gamlss.dist)", y = "Empirical cdf (my function)") + 
  theme(legend.position = "none") 
```

Looks pretty good, no?

# Using the custom distribution functions

To finish out my tests of these functions, I'll demonstrate their use in an actual estimation problem. I'll generate data based on a simple generalized linear model with a single predictor $X$, where the outcome $Y$ follows a double-Poisson distribution conditional on $X$. The data-generating process is:

$$
\begin{aligned}
X &\sim N(0, 1) \\
Y|X &\sim DPO(\mu(X), \theta) \\
\log \mu(X) &= 2 + 0.3 \times X
\end{aligned}
$$
To make things interesting, I'll set the dispersion parameter to $1 / \theta = 0.7$ so that the outcome is _under_-dispersed relative to the Poisson. 

The following code generates a large sample from the data-generating process. To keep things R-centric, I use `gamlss.dist::rDPO` to generate the outcome.

```{r}
set.seed(20230913)
N <- 500
X <- rnorm(N)
mu <- exp(2 + 0.3 * X)
theta_inv <- 0.7
Y <- rDPO(N, mu = mu, sigma = theta_inv)
dat <- data.frame(X = X, Y = Y)
```

Here's what the sample looks like, along with a smoothed regression estimated using a basic cubic spline:

```{r, fig.width = 6, fig.height = 4, fig.retina = 2}
ggplot(dat, aes(X, Y)) + 
  geom_point(alpha = 0.1) + 
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs")) + 
  theme_minimal()
```
Surely the simplest, quickest, and dirtiest way to estimate such a regression is with a generalized linear model, using the "quasi-Poisson" family to allow for non-unit dispersion. In R:

```{r}
quasi_fit <- glm(Y ~ X, family = quasipoisson(link = "log"), data = dat)
summary(quasi_fit)
```
This approach recovers the data-generating parameters pretty well, with a dispersion estimate of `r round(summary(quasi_fit)$dispersion, 3)` compared to the true dispersion parameter of `r theta_inv`. The intercept is under-estimated a little bit, and the slope is over-estimated by a touch.

Now let me fit the same generalized linear model but assuming that the outcome follows a true Poisson distribution (with unit dispersion). I'll fit the model in a Bayesian framework with the `brms` package.

```{r Poisson, cache = TRUE}
Poisson_fit <- 
  brm(
    Y ~ X, family = poisson(link = "log"),
    data = dat, 
    warmup = 500, 
    iter = 2000, 
    chains = 2, 
    cores = 2,
    seed = 20230913
  )

summary(Poisson_fit)
```
This specification recovers the intercept and slope parameters well too, but doesn't provide any estimate of dispersion. 

As an alternative, I'll also fit the model using the negative binomial distribution, which is a generalization of the Poisson that allows for over-dispersion (but not underdispersion):

```{r negbin, cache = TRUE}
negbin_fit <- 
  brm(
    Y ~ X, family = negbinomial(link = "log"),
    data = dat, 
    warmup = 500, 
    iter = 2000, 
    chains = 2, 
    cores = 2,
    seed = 20230913
  )

summary(negbin_fit)
```
The `brms` package implements the negative binomial using the rate parameterization and a log transformation, so the `shape` parameter corresponds to the log of the inverse dispersion. Thus, a large shape parameter (as in the above fit) corresponds to dispersion that is very, very close to one (i.e., close to the Poisson).

Finally, I'll fit the same model but using my custom-built double-Poisson distribution. Following [Paul Buerkner's vignette](https://cran.r-project.org/web/packages/brms/vignettes/brms_customfamilies.html) on using custom distributions, I'll first specify the custom family object for the double-Poisson:
```{r}
double_Poisson <- custom_family(
  "double_Poisson", dpars = c("mu","theta"),
  links = c("log","log"),
  lb = c(0, 0), ub = c(NA, NA),
  type = "int",
  vars = "vint1[n]"
)
```
I set the defaults to use a log-link for the mean (just as with the Poisson and negative binomial families) and a log-link for the inverse-dispersion. 
The argument `vars = "vint1[n]"` is to allow passing a value for the `max_value` argument of 
Next, I'll create an object to add the custom stan code from above into the code created by `brm` for fitting the model:
```{r}
double_Poisson_stanvars <- stanvar(scode = "stancode_qr", block = "function")
```
Now I'm ready to fit the model:
```{r negbin, cache = TRUE}
DPO_fit <- 
  brm(
    Y | vint(1000) ~ X, family = double_Poisson,
    stanvars = double_Poisson_stanvars,
    data = dat, 
    warmup = 500, 
    iter = 2000, 
    chains = 2, 
    cores = 2,
    seed = 20230913
  )

summary(DPO_fit)
```


```{r, fig.width = 10, fig.height = 5}
Yrep_Poisson <- posterior_predict(Poisson_fit, draws = 500)
Poisson_dens <- ppc_dens_overlay(dat$Y, Yrep_Poisson[1:100,])
Poisson_ecdf <- ppc_ecdf_overlay(dat$Y, Yrep_Poisson[1:100,])

Yrep_negbin <- posterior_predict(negbin_fit, draws = 500)
negbin_dens <- ppc_dens_overlay(dat$Y, Yrep_negbin[1:100,])
negbin_ecdf <- ppc_ecdf_overlay(dat$Y, Yrep_negbin[1:100,])
(Poisson_dens + Poisson_ecdf) / (negbin_dens + negbin_ecdf)
```

# Colophon

```{r, include = FALSE, warning = FALSE}
file.remove("DPO-lpmf.stan")
file.remove("DPO-cdf.stan")
file.remove("DPO-quantile.stan")
file.remove("DPO-rng.stan")
```

```{r, echo = FALSE}
sessionInfo()
```


